|

### 2024-03-13 manuscript/manuscript.qmd
manuscript updates

format:
  docx:
    toc: false
  elsevier-pdf:
    journal:
      cite-style: number
      formatting: preprint
      model: 3p
      layout: twocolumn
      #cite-style: number
    include-in-header: 
      text: |
        \newpageafter{author}
  html:
    page-layout: article
    toc: true
    toc-depth: 1
    toc-expand: 2
    toc-title: Table of Contents
    code-tools:
      source: true
      toggle: false
      caption: none
    grid:
      sidebar-width: 350px
      margin-width: 20px

#fontfamily: libertinus
#fontsize: 20pt
#colorlinks: true
#format-links: [html, pdf]
#execute:
  #echo: false
#bibliography: "references.bib"
#biblio-style: unsrt

@@

keywords:
- ABCD Study
- longitudinal analyses
- development
---

@@

\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (Volkow et al. 2018). At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes. Central to achieving these goals is the commitment of the ABCD Study® and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods, by espousing practices that increase access, integrity, and reproducibility of scientific research. In this sense, the ABCD Study is a collaboration with the larger research community.

@@

![Types of Stability and Change](./figures/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}

### Use of appropriate longitudinal models
There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2011; @hoffman2015; @littlefield2021].

First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, [e.g., @curran2011]). Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015; latent curve models with structured residuals [LCM-SR], @curran2014). It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

@@

Given that longitudinal data involve multiple sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data. An effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014]. Common effect size metrics include the correlation r between two variables and the standardized difference between two means, Cohen’s *d* [@cohen1988]. An extensive discussion of effect sizes and their relevance for ABCD is given in @dick2021. Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered (e.g., @morris2002]. @wang2019 note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic [see @wang2019, for more details]. Thus, obtaining effect size metrics based on standardized estimates that are relatively simple in cross-sectional data (such as r) becomes more complex in the context of prospective longitudinal data. @feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis [see @feingold2009, for more details]. Given this issue, there have been various proposals for adjusting effect size measures in repeated assessments. @feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (i.e., quadratic) growth factors. @morris2002 review various equations for effect size calculations relevant to combining estimates in meta-analysis with repeated measures and independent-groups designs. Other approaches to quantifying effect sizes longitudinally may be based on standardized estimates from models that more optimally disentangle between- and within-person sources of variance. As an example, within a random-intercept cross-lagged panel model (RI-CLPM) framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.

@@

[![Longitudinal Data Analysis Structural Diagram](./figures/LDAStructure2.png)](https://centerstat.org/resources/)

@@

![Longitudinal Models/Data Structures](./figures/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"} <br>

#### Traditional Models

Traditional methods for longitudinal analysis primarily focus on modeling mean-level change, and how these changes may differ across groups or levels of some other variable. For example, is there a difference in average internalizing symptoms obtained across multiple assessments between boys and girls? Longitudinal models that focus on mean-level change are also referred to as marginal models and examples of specific methods include repeated measures ANOVA (and MANOVA), ANCOVA, and Generalized Estimating Equations (GEEs). Mean-level change models are commonly used when data is only available from 2 measurement occasions. For example, computing a difference score (e.g., mean internalizing scores at time 2 - mean internalizing scores at time 1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups. Additionally, the longitudinal signed-rank test, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data. Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level [@castro-schilo2018].

@@

#### Modern GLM Extensions

Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability. Generalized Estimating Equations (GEE), Linear Mixed Models (LMM), Generalized Linear Mixed Models (GLMM), and Autoregressive Cross-Lagged Panel Models (ARCL) are examples of such contemporary techniques. GEE, an extension of Generalized Linear Models, combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects. For example, @vandijk2021 used GEE to obtain relative risks for psychiatric diagnoses among children in the ABCD Study with a family history of depression and used the ABCD Study sampling weights to generalize prevalence rates among 9 and 10 year-olds across the US. LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures. Within the ABCD Study, reseachers may want to consider nesting by individual, family (i.e., siblings or twins), school or district, and/or site. GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use [e.g., @martz2022] screen media use [@lees2020], and microstructure of the brain [@palmer2022]. Finally, ARCL models are used to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects [although, ARCL models are relatively less useful for teasing apart between-person and within-person sources of variances; see @curran2021].

@@

#### SEM

Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error. SEM techniques can be categorized into variable-centered, person-centered, and hybrid approaches, each with unique strengths and limitations. The choice of method depends on the research question, data structure, and underlying assumptions.

@@

#### Advanced SEM

Advanced structural equation modeling (SEM) approaches, such as the RI-CLPM and LCM-SR models, have emerged to address more complex research questions and data structures in longitudinal analysis. These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.

@@

![LDA Table](./figures/Tabletemp.png)

# Discussion

@@

{{< pagebreak >}}

# Part V: References

>>>>

@@

index-terms: 
  - Longitudinal Analysis
  - ABCD Study
  - Development

format:
  ieee-typst: default
  html:
    page-layout: article
    toc: true
    toc-depth: 1
    toc-expand: 2
    toc-title: Table of Contents
    code-tools:
      source: true
      toggle: false
      caption: none
    grid:
      sidebar-width: 350px
      margin-width: 20px
bibliography: references.bib

---

@@

[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (Volkow et al. 2018). At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes. Central to achieving these goals is the commitment of the ABCD Study® and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods, by espousing practices that increase access, integrity, and reproducibility of scientific research. In this sense, the ABCD Study is a collaboration with the larger research community.

@@

[!Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}

### Use of appropriate longitudinal models
There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2011; @hoffman2015; @littlefield2021]. First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, [e.g., @curran2011]). Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015; latent curve models with structured residuals [LCM-SR], @curran2014). It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context (see @littlefield2022, for further discussion).

@@

Given that longitudinal data involve multiple sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data. An effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014]. Common effect size metrics include the correlation r between two variables and the standardized difference between two means, Cohen’s *d* [@cohen1988]. An extensive discussion of effect sizes and their relevance for ABCD is given in @dick2021. Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered (e.g., @morris2002). @wang2019 note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic (see @wang2019, for more details). Thus, obtaining effect size metrics based on standardized estimates that are relatively simple in cross-sectional data (such as r) becomes more complex in the context of prospective longitudinal data. @feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis (see @feingold2009, for more details). Given this issue, there have been various proposals for adjusting effect size measures in repeated assessments. @feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (i.e., quadratic) growth factors. @morris2002 review various equations for effect size calculations relevant to combining estimates in meta-analysis with repeated measures and independent-groups designs. Other approaches to quantifying effect sizes longitudinally may be based on standardized estimates from models that more optimally disentangle between- and within-person sources of variance. As an example, within a random-intercept cross-lagged panel model (RI-CLPM) framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.

@@

[![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png)](https://centerstat.org/resources/)

@@

[!Longitudinal Models/Data Structures](img/Mindmap1copy.png)]{#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}

[![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png)](https://centerstat.org/resources/)
<br>

#### Traditional Models
Traditional methods for longitudinal analysis primarily focus on modeling mean-level change, and how these changes may differ across groups or levels of some other variable. For example, is there a difference in average internalizing symptoms obtained across multiple assessments between boys and girls? Longitudinal models that focus on mean-level change are also referred to as marginal models and examples of specific methods include repeated measures ANOVA (and MANOVA), ANCOVA, and Generalized Estimating Equations (GEEs). Mean-level change models are commonly used when data is only available from 2 measurement occasions. For example, computing a difference score (e.g., mean internalizing scores at time 2 - mean internalizing scores at time 1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups. Additionally, the longitudinal signed-rank test, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data. Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level [@castro-schilo2018].

@@

#### Modern GLM Extensions
Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability. Generalized Estimating Equations (GEE), Linear Mixed Models (LMM), Generalized Linear Mixed Models (GLMM), and Autoregressive Cross-Lagged Panel Models (ARCL) are examples of such contemporary techniques. GEE, an extension of Generalized Linear Models, combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects. For example, @vandijk2021 used GEE to obtain relative risks for psychiatric diagnoses among children in the ABCD Study with a family history of depression and used the ABCD Study sampling weights to generalize prevalence rates among 9 and 10 year-olds across the US. LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures. Within the ABCD Study, reseachers may want to consider nesting by individual, family (i.e., siblings or twins), school or district, and/or site. GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use [e.g., @martz2022] screen media use [@lees2020], and microstructure of the brain [@palmer2022]. Finally, ARCL models are used to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects [although, ARCL models are relatively less useful for teasing apart between-person and within-person sources of variances; see @curran2021].

@@

#### SEM
Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error. SEM techniques can be categorized into variable-centered, person-centered, and hybrid approaches, each with unique strengths and limitations. The choice of method depends on the research question, data structure, and underlying assumptions.

@@

#### Advanced SEM
Advanced structural equation modeling (SEM) approaches, such as the RI-CLPM and LCM-SR models, have emerged to address more complex research questions and data structures in longitudinal analysis. These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.

@@

<!-- ![LDA Table](img/Tabletemp.png) -->

# Discussion
\label{sec:headings}

@@
